Skeleton World - the (example) game.



Decisions:

Scale: Simplicity. I want a simple game, scaled back where I can. \sidenote{This is becuase it is an example, so the less I have to write the better as it will make the decisions and strucutre of the example clear. There is also a thematic unity between playing as skeletons and a bare-bones game}

Key Imagery: A sword weilding skeleton emerging from the ground; cartoon skeleton's in a danse macarbe; a burning skeleton throwing fireballs, a knight in armour made from hundred of ribs; a mishapen skeletal beast with too many limbs and joints that bend backwards, a screaming skull floating, a skeleton in a suit backing away in horror as plants consume another.

Community vs Party: Adversial. I want to chase out the bleak loneliness of death and the grimdarkness of being a skeleton, trapped in a timeless world, being slowly ground down against each other. \sidenote{this was a key decision. A game about a special forces unit in a Necromancer's Army would be very different. And possibly more fun?} 

Key Theme: Change. The world and enviroment are unstable and shifting. \sidenote{this was chosen partly be default as non of my other examples in the appendix seemed to fit.}
Secondary Theme: Scarcity. The world is basically a wasteland with only a few key items belonging to the players. This combined with Change suggest dunes blowing. What makes the sand? the ground down bones of ancient skeletons of course!
Palette: Emphasis on bones and the limited things the skeletons bring through from the living world.

Statistics: Lets go for a simple variant on mind, body, spirit as BONES, MEMORY, GLUE
Experience: Let's go for a simple method. At the start of each session, each player gives a brief recap of what they did and marks a new move. No experience is tracked.
Playbooks: The Key imagery suggests a few already. Two points spread across three stats gives six unique statlines, lets use them as the basis. Let's give each a unique glue to add a bit of imagery to split them up, and lets assign each playbook two of St John Ross's 12 RPG archetypes, it will give the playbooks a bit more depth.

BONES 0, MEMORY 2, GLUE 0  Splint+clothes  		Persuading					Gadegeeter
BONES 1, MEMORY 1, GLUE 0  Beast+wires			Wilderness survival		Intrusion
BONES 0, MEMORY 1, GLUE 1  Mage+fire				Communication+Protocol	Scholarship
BONES 2, MEMORY 0, GLUE 0  Knuckles+scrimshaw  Athletics					Drive/Ride
BONES 1, MEMORY 0, GLUE 1  Glom+guts				Medical						Combat
BONES 0, MEMORY 0, GLUE 2  Skull+plant				Detection					Weird


Basic Moves

Struggle
When you struggle for control of something, roll+Bones.\sidenote{So many decisions here. First is that since there are only three stats I want to limit it to three thematic ways to take action, thwe trigger here covers combat, but also argument, intimidation and person versus an uncaring universe. Second is that struggling for control is dangerous, so damage is delivered on nearly all options. Third is that, for the basic moves at least, I'll prescribe a 6- result to help the players expectations stay in genre. Fourth is the introduction of a new mechanic, just adding a little extra spice to the dice roll. With only three moves I don't want them to feel too samey in their results.}
On a 10+ exchange damage and take control of the thing
On a 7-9 exchange damage, take control of the thing and give up a position, something valued or an honest answer to your attacker. \sidenote{you may feel that an `honest answer' is the obvious least-cost choice, but it acts as a good way for the MC to inject loaded questions to draw out characters. The other options are protection for players who hate being put on the spot like that.}
On a 6- the result is as 7-9 but you don't get control.
If you also roll a double on any result, don't take damage.

Unleash \sidenote{can't have a limbo based game without good and bad memories showing up, and this is a great place for the change theme to be contrasted. Name reduction is a type of HP/manna clock and seemed a thematic way to avoid memory spamming. At this point in the design I'm not sure if I will allow a way to recharge your name with fresh letters, or if the pacing should be to allow one use per session (on average) for a six session campaign. }
When you unleash a memory to try and shape something out of the dunes Roll+Memory
On a 10+ it is there and will not change, ask the MC how many letters to reduce your name by
On a 7-9 as 10+ but it is weak, temporary or wild
On a 6- something bad is unleashed instead. It will haunt you.

Take Damage \sidenote{ a little bit of randomisation which means that weapon descriptions can have fixed damage. It is primarly to cover the genre image of a skeleton pulling itself back together from a thousand pieces. The suitable repair is supposed to generate a hook/pressure for the player character to take action.}
When you take damage, even zero damage after armour, Roll+Glue
On a 10+ it's not so bad, take the damage but reduce it by one or take take the 7-9 option.
On a 7-9 take the damage OR ignore damage but reduce Glue to zero until you suitably repair it (see playbook)
On a 6- take the damage. Your armour fails to do anything.  OR ignore damage but reduce Glue to zero until you suitably repair it (see playbook)

Decision: No move for helping/hindering another player. It's a struggle out there. 

So that is three stats that have led to three basic moves (and threee little extra mechanisims - rolling doubles, reducing name length, repairing Glue). That is ok for now, but I hope to tighten that up with a bit of playtesting.


Playbook: Splint
BONES 0, MEMORY 2, GLUE 0  
Name - six letters long
Glue - clothes . Repair by sewing, patching (if you can find the needle and thread) or replacing items of clothing (if you can take them). 		
Your job in life is mixed up in your identity, such that some of your bones are replaced by items symbolic to you. 
Start Move: Start with two implements (choose anything from the equipment list). Implements are always available, but cannot be removed or given to others. Say what body part they replace.
[] Junkyard: detail another implement that is now part of you
[] Substitute: `consume' a suitable item to heal three damage
[] Donate (needs Substitute): Spend three health to give anyone a previouly subsituted item
[] My Uniform: Increase Memory by +1 permenantly
[] Strangely Dishonest: When you lie to their face Roll+Memory. 
On a 10+ they buy it, and it may even change the world. 
On a 7-9 they buy it for now.
[] Presence: When you unleash a memory, you can say who, if present, is unable to look away. 
[] Demagouge: When you try to lead a volatile crowd Roll+Memory
On a 10+ if you meet their needs, they'll meet yours
On a 7-9 as 10+, but strictly in that order!

Equipment list: Wrench, till recipet roll; typewriter; chef knife; teapot; stethoscope; pens pens pens; mop; hard hat and boots; camera; measuring tape; material shears; every car key; globe


Playbook: Beast
BONES 1, MEMORY 1, GLUE 0  
Name: Three letters long
Glue: Thin brass wires. Repair by scavenging the wires from broken or moving dune monsters.
You've been here a long time, if time even passes here. There are many that have entered here before the `humans', and you are cunning enough to have survived.

[] Start with one of: Big Teeth (+3 damage), Huge size (+2 armour), tattered wings (limited flight)
[] Lamarkian: Take another starting thing. Duplication allowed.
[] Reputation: Add four more letters to your name
[] Cunning: When you hide in the dunes Roll+Mem
On a 10+ you are undetectable until you decide to act
On a 7-9 you are invisible to common appearances
[] Shoving: When you struggle with another beast ritually over territory or similar, on a 6- you won't exchange damage.
%[] Pounce: When you commit to pin someone down, roll+Bones
%On a 10+ they are trapped
%On a 7-9 you take damage but they are trapped.
%On a doubles, don't take damage. \sidenote{this needs to be added to avoid the move being worse then straightforward struggle. It dosen't add anything to the move. Perhaps the double mechanic should go, or Pounce is simply not different enough from Struggle to justify existence}
[] Cage Breaker: When you pick a lock with your long wire tounge, Roll+Glue
On a 10+ it is delicate and quick
On a 7-9 it is neither, but you get the lock open
[] Call of the Wild: When faced with an constructed obstacle you break or bypass Roll+Bone
On a 10+ you succeed or the MC suggests an alternative good route
On a 7-9 you succeed. Enjoy your new situation! (ask the MC)




Playbook: Mage
BONES 0, MEMORY 1, GLUE 1  
Name: five letters long
Glue: coloured fire. Repair by lighting and building a small fire, and bathing in it.
Obssessed or just passionate, you can't see why everyone dosen't agree with you.
Start with Projection: You scoop and throw a fireball of desire dealing 2 damage.
[] Griefer: You pull a blinding cloud of ash and embers out of the ground around you.
[] Furnace When you crush someone into your rib hollow Roll+Glue
On a 10+ do 10 damage to them
On a 7-9 as 10+ but you reduce Glue to zero until repaired
[] Hatred: When you spit your fire on someone Roll+Glue:
On a 10+ it clings and continue to burn and damage them until you speak to them
On a 7-9 as 10+ but strangely, although it burns it doesn't actually damage them.
[] Scholar: When you study something carefully Roll+Memory
On a 10+ you spot something no-one else would
On a  7-9 you have an idea about it
[] Enlighten: When you are stifled by rules Roll+Glue
On a 10+ you guess who to challenge to unmake the system
On a 7-9 you guess where the power lies
[] Flare: When you close your eyes you can appear as a small fiery image to anyone you have previously damaged. This power may reach over distance and through walls.

Playbook: Knuckles
BONES 2, MEMORY 0, GLUE 0  
Name = eight letters
Fire: Scrimshaw. Repair by having an artist add scrimshaw to replacement clean bones.
The simplest approach is often the most direct.
Start with: Brawler:Even unarmed you still do 3 damage
[] Bullish: When you hit someone hard you can choose to knock them backwards
[] Naturally Spiky: Reduce damage taken by 1. This does no stack with armour.
[] Athlete: When you push to do something others can't Roll+Bones
On 10+ you do it.
On 7-9 you do it but take 1 damage.
[] Artistic: As long as you have a working hand, you can record your feats in scrimshaw (counts for repair glue). Otherwise you must find an artist.
[] Domination: When you wrestle a lage dune monster into submission Roll+Bones \sidenote{struggle theme again -  no easy loyal animal companions here!}
On a 10+ it accepts you. Treat it well and it will let you ride it.
On a 7-9 as 10+ but it dislikes you and will use every opportunity to escape.
[] Martial Dancer: When you prepare to fight Roll+Mem
On 10+ hold 3,
On 7-9 hold 1,
During the fight, spend hold one for one to - Move unstoppably; break a grip; block an attack

Playbook: Glom
BONES 1, MEMORY 0, GLUE 1  
Name: six letters
Glue: Guts. Repair by consuming fresh meat or funeral foods.
You always blended in well, and bones are bones are bones right?

Start with: Amalgamate: When you gut wrap sufficent new bones into any skellington Roll+ Glue: 
On a 10+ heal 3 or heal someone else 3 
On a 7-9 as 10+ but reduce glue to zero until repaired
[] Gall Stones: You develop a pot belly stuffed with bones and sand to digest even old bones. Consuming a full skellington is enough for you to repair your glue.
[] Blemeye: Your skull moves down inside your ribcage. Increase you base armour by 2. More armour stacks with this.
[] Cage and Club: One arm ends in a ribcage sheild, the other in another femur. You always count as armed with a Large Club, and the sheild increases your base armour by 1. 
[] Gut Lassoo: Needs Gall Stones or Cage and Club. When you throw (or spit) a lassoo made of your guts, roll+Glue.
On a 10+ you get them. You can unbalance, disarm them or draw them closer.
On a 7-9 you get them but it's awkard. Choose as 10+ but then they choose an option to do to you.
[] Double Jointed: Needs Cage and Club or Blemeye: You gain a second pair of arms. If this means you have four hands this allows you to duel weild Great weapons. Otherwise it just means you have hands free to carry things.
[] 


Playbook: Skull
BONES 0, MEMORY 0, GLUE 2  
Name: eight letters 
Glue: Plant. Repair by filling skull with water and staying in the sun for a short time.
In a world of white sands, what is left of you brings the true death. 

Start with: Manifest: When you need to be more then a slowly floating skull Roll+Glue
On a 10+ you manifest a knotted viney body, as strong and fast as any skellington.
On a 7-9 Choose one: you are small, limbless or slow
[] Rooted: When you appear to remain motionless, you may push a limb out of the dust three body lengths away.
[] Infectious growth. When you deal damage and roll doubles you can choose to ignore their damage or to infect them. Infected take one damage each turn ignoring armour.
[] Tendrils: No-one can sneak up of you. Your attacks can entangle.
[] Blossom: When you exert power through an infected, they blossom with large flowers. They become the focus of the attention of anyone around.
[] Detection: 
[] Weird:
				Detection					Weird


